---
title: "Programming with models: Writing nimbleFunctions to implement algorithms"
subtitle: "enviBayes 2021 tutorial"
author: "NIMBLE Development Team"
output:
  html_document:
    code_folding: show
---


```{r chunksetup, include=FALSE} 
# include any code here you don't want to show up in the document,
# e.g. package and dataset loading
library(methods)  # otherwise new() not being found
library(nimble)
```

# Introduction

*nimbleFunctions* are at the heart of NIMBLE. They are the way that algorithms are implemented. They can also be used for

 - user-defined BUGS distributions (already seen),
 - user-defined BUGS functions (not seen, but similar to user-defined distributions)
 - user-defined MCMC samplers (coming soon), and
 - compiling parts of R (not seen), without reference to a model.

But their main purpose is providing a way for developers to implement algorithms.

# Why develop an algorithm in NIMBLE?

 - You get flexible hierarchical modeling infrastructure "for free".
 - The NIMBLE system allows you to program essentially arbitrary computations.
 - You can combine your method with or build on top of  ones already available in NIMBLE.
 - You can distribute your method as a separate package that uses NIMBLE.


# Components of a nimbleFunction

NIMBLE uses the concept of *two-stage evaluation* from computer science to run a model-specific algorithm based on model-generic algorithm code. The first stage of evaluation specializes the algorithm to the model of interest via *setup* code. The second stage runs the algorithm via *run* code. 

Thus, a nimbleFunction has two parts:

 - setup code: used to tailor the algorithm to a particular model structure. Often this involves determining dependencies amongst nodes in the model and setting up storage using *modelValues*
 - run code: the guts of the algorithm, written generically so it will apply to any (appropriate) model

Setup code is written as a R function, using R code, usually including NIMBLE's special functions for querying the model structure (see the [module on querying model structure](10_model_structure.html)). 

Run code is written using the NIMBLE *domain-specific language* (DSL). While this is formally a language distinct from R, you can just think of it as a subset of R, enhanced with some functions for operating on the model (see the [module on operating a model](9_operating_model.html)). 

# Some syntax for nimbleFunctions

Here are some of the functions you may use in the run function of a nimbleFunction:

 - *returnType*, e.g., ```returnType(double(1))``` for a vector of reals
 - *length*, e.g., ```length(x)``` to determine the length of a run-time argument *x*
 - *numeric*, *matrix* and *array* e.g., ```result <- numeric(n, init = 1.0)``` to create a vector of reals called *result* initialized with values of 1.0
 - model member functions *calculate*, *simulate*, *getLogProb*, *calculateDiff* and *getParam* to manipulate the model
 - direct access to nodes or variables in a model using typical R syntax, e.g., ```model[[myNode]] <- rnorm(1)```
 - *values()* and *copy()* (or, equivalently, *nimCopy*) to copy values
 - *print()* and *cat()*
 - basic math, including vectorized math and some linear algebra
 - random number generation functions, e.g., ```rnorm(1, 100, 5)``` 
 - calling out to arbitrary R or C/C++ code with *nimbleRcall()* and *nimbleExternalCall()*
 - *nimbleList* data structures.


Section IV of the NIMBLE User Manual describes the syntax for *run* code in detail, including lots of neat functionality such as using nested nimbleFunctions and having multiple run-time functions (i.e., class methods) as part of a nimbleFunction. We'll see more of that in future modules.

This [tips and tricks page](https://github.com/nimble-dev/nimble/wiki/Tips-for-avoiding-errors-in-nimbleFunction-programming-and-compilation-%28DSL%29) details some syntax that does NOT work due to some clunkiness in our compilation system. 

nimbleFunctions use **pass-by-reference** not R style **pass-by-value**, so be careful about modifying an object and then using it elsewhere.
  

# A basic example: empirical Bayes / maximum marginal likelihood

Let's consider how we would optimize the parameters in a model using a nimbleFunction. Basically, we'll just construct an objective function that we can then pass to R's *optim* function to do the actual numerical optimization. (NIMBLE also has an `optim()` that you can use within a nimbleFunction.)

This amounts to setting things up to find the posterior mode of a model; this is generally a reasonable thing to do only for models with a small number of parameters and without hierarchical structure. That sounds restrictive, but if you can marginalize out the latent process values, then we're doing empirical Bayes.

# A nimbleFunction for the litters marginalized model

```{r, litters-objective, eval}
objective <- nimbleFunction(
    setup = function(model) {
          # ordinarily we would do stuff here, but in this case
          # we only need make the nimbleFunction aware of the model
          },
    run = function(par = double(1)) {
        returnType(double(0))
        ## Assignment into model or variables created in setup code
        ## requires global assignment operator, <<-
        model[['a']] <<- exp(par[1:2])
        model[['b']] <<- exp(par[3:4])
        ans <- model$calculate()
        return(ans)
    }
)
```

This is actually a nimbleFunction *generator* -- we can't run it yet -- we need to create a specialized instance of the nimbleFunction that is tailored for some model, in our case the marginalized litters model. 

# Specializing the nimbleFunction to the model

First let's build the marginalized litters model again.

```{r, rebuild, results='hide', message=FALSE}
source("chunks_litters_marginal.R")
```

Now we specialize the algorithm to the specific model object.

```{r, litters-specialized}
rObjective <- objective(littersMargModel)
cObjective <- compileNimble(rObjective, project = littersMargModel)
```

Now let's try using it.

```{r, litters-optimize}
system.time(optR <- optim(log(rep(1,4)), rObjective$run,
                 control = list(fnscale = -1)))
system.time(optC <- optim(log(rep(1,4)), cObjective$run,
                 control = list(fnscale = -1)))
optR
optC
exp(optC$par)
```

# Writing generic functions

Let's look back at our nimbleFunction objective function. What stops it from being usable on any model?

```{r, nf-generic}
objective <- nimbleFunction(
    setup = function(model, target) {
          ## We'll start putting stuff here soon, I promise!
          },
    run = function(par = double(1)) {
        returnType(double(0))
        values(model, target) <<- exp(par)
        ans <- model$calculate()
        return(ans)
    }
)
```

# Writing generic functions - querying model structure

Calculating the density for all model nodes is not necessary for this optimization, since we're doing maximum likelihood rather than maximum a posteriori (MAP) estimation. In particular we don't need to calculate the prior density values.

```{r, nf-generic2}
objective <- nimbleFunction(
    setup = function(model, target) {
        ## We only need to calculate the likelihood, not the full model density.  
        calcNodes <- model$getDependencies(target, self = FALSE)
        cat("Maximixing w.r.t. ", paste(calcNodes, collapse = ','), ".\n", sep = "")
    },
    run = function(par = double(1)) {
        returnType(double(0))
        values(model, target) <<- exp(par)
        ans <- model$calculate(calcNodes)
        return(ans)
    }
)
```


```{r, litters-optimize2}
rObjective <- objective(littersMargModel, c('a', 'b')) ## or c('a[1]','a[2]','b[1]','b[2]')
cObjective <- compileNimble(rObjective, project = littersMargModel)
optC <- optim(log(rep(1,4)), cObjective$run, control = list(fnscale = -1))
optC
```

